Chemical tagging: \\ Identification of stellar phase-space structures \\ purely by chemical-abundance similarity

### 2016-01-14 papers/clusters.tex
intro text zeroth draft

% to-do
% - finish writing MKN's introduction
% - write data and cannon section
% - write method and results section
% - write discussion section
% - put in references
% - cite codes!
% - put in acknowledgement
% - forward to Freeman, others?
% - submit to journal and arXiv

@@

\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\gaiaeso}{\project{Gaia--\acronym{ESO}}}
\newcommand{\galah}{\project{\acronym{GALAH}}}
\newcommand{\thecannon}{\project{The~Cannon}}
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\eg}{\foreign{e.\,g.}}
\newcommand{\etal}{\foreign{et\,al.}}

@@

Stellar spectroscopic surveys now have the resolution, signal-to-noise,
wavelength coverage, and sample sizes to deliver chemical tags.
The issue at present is measurement precision (and accuracy).
The current precision on abundance measurements is not high enough
(for example, \citealt{Ting2015, Martel2015}).
That said, it is clear that the \emph{data} are precise enough;
there is enough signal-to-noise at the relevant locations in spectrum
space to deliver high-precision tags.
Our view is that the traditional methods for making abundance
measurements just aren't mature enough:
There are issues with the physical models, and the data-analysis
methods do not make use of all of the information in the data sets.
Improving the models and exploiting the entire information
content in the data is critical if we are going to deliver useful
chemical tags.

One note on accuracy and precision:

In principle, the

These have not successfully identified components of the galaxy's
assembly that had distinct formation sites (\eg, Ting et al., 2015, Mitschang. et al. 2013,
Blanco-Cuaresma2015).

Blanco-Cuaresma et al., 2015 demonstrate that there is a large degree
of overlap when exploring the chemical space of known open clusters
which challenges chemically tagging and separating them, but also note
that there is room for improvement if more elements were included in
their analysis and models improved.
Ting et al., 2015 presented an algorithm to detect groups of
chemically homogeneous stars which they reported as challenging and
with a low signal to noise in the APOGEE dataset.
Mitschang et al., 2013 propose that chemical tagging may be able to
identify coeval groups of stars but these are not necessarily from a
distinct origin of formation.

The by-hand study of solar twins, for a handful of stars, has shown
however that the chemical information in some elements offers powerful
signatures for chemical tagging (Melendez2014, Jofre2015) and PCA
approach of Ting et al., 2014 shows there are many independent
dimensions of information contained in the multi abundance phase space
being delivered by the large spectroscopic surveys (e.g. Ting et al.,
2014 report a dimensionality of 7 for GALAH survey).
These approaches are important to establish which elements should be
targeted in surveys.

DWH:  We successfully achieve chemical tagging and we do so because we are
working with a higher precision that previously and exploit the full
information in the data....

DWH: With \thecannon\ we achieve both higher precision and exploit the full
information content of the data.....cite Casey et al., 2016 Ness et
al., 2016 in prep = do you want me to write things about this...

DWH: We do this with the APOGEE survey which covers a much wider
spatial extent but GALAH will be much more powerful as they will
report 29 elemental abundances which carry a larger quantity of
nucleosynthetic information...

DWH: Open clusters look one-metal (CITE); globulars look peculiar
(CITE); but no-one has ever gone the other way!...

DWH: Fast intro to \apogee.

DWH: Fast intro to \thecannon.

DWH: BASH the \apogee\ discontiguous field positioning.

DWH: Think of this as a hypothesis test.

>>>>

% to-do
% - write data and cannon section
% - write method and results section
% - write discussion section
% - put in references
% - cite codes!
% - put in acknowledgement
% - forward to Freeman, others?
% - submit to journal and arXiv

@@

% text shit
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\gaiaeso}{\project{Gaia--\acronym{ESO}}}
\newcommand{\galah}{\project{\acronym{GALAH}}}
\newcommand{\thecannon}{\project{The~Cannon}}
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\eg}{\foreign{e.\,g.}}
\newcommand{\etal}{\foreign{et\,al.}}
\newcommand{\documentname}{\textsl{Article}}

% math shit
\newcommand{\teff}{T_{\mathrm{eff}}}
\newcommand{\logg}{\log g}

@@

Stellar spectroscopic surveys now have the resolution, signal-to-noise,
wavelength coverage, and sample sizes to deliver chemical tags.
There are, however, two big issues.
The first is that the physical assumptions behind the idea may require
refinement:
There may be chemical-abundance overlaps among open clusters
\citep{blancocuaresma}; coeval groups of stars may have similar tags
but different birth places \citep{mitschang}, and the
chemical-abundance space might be low in dimensionality \citep{what}.
On the other hand, precise studies of stellar twins \citep{melendez, jofre}
and open clusters \citep{bovy} indicates that chemical abundances ought to be
very precisely similar for co-eval stars.

The second issue for chemical tagging---and the one we address
here---is measurement precision (and accuracy).
The current precision on abundance measurements is not high enough
(for example, \citealt{Ting2015, Martel2015}).
That said, it is clear that the \emph{data} are precise enough;
there is enough signal-to-noise at the relevant locations in spectrum
space to deliver high-precision tags.
Our view is that the traditional methods for making abundance
measurements just aren't mature enough:
There are issues with the physical models, and the data-analysis
methods do not make use of all of the information in the data sets.
Improving the models and exploiting the entire information
content in the data is critical if we are going to deliver useful
chemical tags.

Our contribution in this space has been to develop
\thecannon\ \citep{cannon, ages}, which is a data-driven model for
stellar spectra.
This model can deliver stellar parameters and chemical abundances to
stars, making use of every pixel of every stellar spectrum (that is,
all the information in the data) but making no use at all of physical
models of stars.
It relies only on there being training data---\emph{some} stars for
which parameters and abundances are known and believed.
In companion papers \citep{casey16, ness16} we show that
\thecannon\ can deliver 15 to 19 abundances for stars in the
\apogee\ survey, at precisions higher than even the training data on
which the model is trained.
We say more about this below.
We will show here that \thecannon\ improves chemical abundance measurements
to the point that \emph{chemical tagging is now possible.}
We have thereby effectively saved the collective asses of all the
collaborators in all of the above-referenced large spectroscopic
surveys.

One note on \emph{accuracy} and \emph{precision:}
In principle, the problem of chemical tagging does not require
absolute accuracy for chemical-abundance measurements.
We only require that we can precisely see that two stars are similar
in their abundances; even if we are wrong about the absolute values of
those abundances.
This point might make it seem like we don't care that our models are
wrong, so long as they are \emph{consistent.}
However, this is a bit misleading:
For chemical tagging to succeed, we need stars with different
atmospheric parameters $(\teff, \logg)$ but the same chemical
abundances to be assigned the same position in chemical-abundance
space.
That doesn't require overall accuracy, but it requires that the models
have the right dependencies on atmospheric parameters such that the
wrongness in abundance space is consistent across the H-R diagram.
That is, we need a substantial amount of accuracy to achieve our goals.

In what follows, we are going to use \apogee\ \acronym{DR12} data, in
which we can measure 15 element abundances (C, N, O, Na, Mg, Al, Si,
S, K, Ca, Ti, V, Mn, Fe, Ni).
The \apogee\ data set covers a huge radial extent and---because the
data are taken in the infrared---explores the thin disk.
However, it has the disadvantage that it's spatial coverage is swiss
cheese, which makes it hard to see, within the data set, linear or
extended stellar structures.
In many ways, \galah\ will deliver improvements, both because it will
have more abundances (possibly 29) and contiguous sky coverage.

One final note:
We think of this \documentname\ as performing a hypothesis test.
We know that the stellar members of open and globular clusters---stars
that are identified by being close in phase space---contain highly
informative abundance information that identifies them also in
chemical-abundance space.
Does this work the other way around?
Are groups of stars that are identified as being close in abundance
space---collections of stars with near-identical chemical tags---also
found to be close in phase space?
If the answer to this question is ``yes'' then chemical tagging is
\emph{on.}
